Light and life form interact to shape stomatal ratio among British angiosperms$^1$

### 2017-07-12 ms/ms.Rnw
added ability to knit ms

<<Import objects, echo = FALSE>>=

  source("r/header.R")

  import2ms()

@@

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_lf-light.pdf}}
\caption{Lifeforms have different tolerances for sun and shade among British angiosperms. Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different Raunki\ae r life forms. Height of the bars indicate the raw proportion of species in each bin for that lifeform. The sample size for each lifeform is listed next in parentheses. The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.} 
\label{fig:lf-light}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_SRmultReg.pdf}}
\caption{The effect of light on stomatal ratio depends on Raunki\ae r life form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in therophytes, hemicryptophytes, and chamaephytes but not geophytes and phanerophytes. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 1000 parametric bootstrap samples. Numbers in parentheses next to Raunki\ae r life form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visial clarity.} 
\label{fig:SRmultReg}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{Figures/Figure_SD-light.pdf}}
\caption{Light-mediated evolution of stomatal ratio is mostly driven by increased adaxial (`upper') stomatal density (Panel A), whereas abaxial (`lower') stomatal density (Panel B) is similar across Ellenberg light indicator values (\el~$x$-axis). The violin plot shows stomatal density ($y$-axis, log-scale) as a function of \el. The width of the grey polygons indicates the density of data. Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles. Points outside the polygons are statistical outliers.} 
\label{fig:SD-light}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics{Figures/FigureS_violin.pdf}}
\caption{Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface (high $\mathrm{SD_{propAd}}$). The violin plot shows stomatal ratio as a function of Raunki\ae r lifeform. The width of the grey polygons indicates the density of data. Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles. Sample sizes per lifeform in the dataset are given above the upper plot margin. $\mathrm{SD_{ad}}$ and $\mathrm{SD_{total}}$ stand for the stomatal density on the adaxial surface and the total leaf surface (adaxial plus abaxial), respectively.}
\label{fig:violin}
\end{figure}

>>>>

@@

<<Import-objects, echo = FALSE>>=

  setwd("..")
  source("r/header.R")

  import2ms()
  setwd("ms")

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_lf-light.pdf}}
\caption{Lifeforms have different tolerances for sun and shade among British angiosperms. Each panel is the distribution of Ellenberg light indicator values on an integer scale of 1-9 for different Raunki\ae r life forms. Height of the bars indicate the raw proportion of species in each bin for that lifeform. The sample size for each lifeform is listed next in parentheses. The mean (open circle) and 95\% confidence intervals (black line) around the mean Ellenberg light indicator value for each lifeform based on phylogenetic regression are above the histogram.} 
\label{fig:lf-light}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SRmultReg.pdf}}
\caption{The effect of light on stomatal ratio depends on Raunki\ae r life form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in therophytes, hemicryptophytes, and chamaephytes but not geophytes and phanerophytes. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 1000 parametric bootstrap samples. Numbers in parentheses next to Raunki\ae r life form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visial clarity.} 
\label{fig:SRmultReg}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SD-light.pdf}}
\caption{Light-mediated evolution of stomatal ratio is mostly driven by increased adaxial (`upper') stomatal density (Panel A), whereas abaxial (`lower') stomatal density (Panel B) is similar across Ellenberg light indicator values (\el~$x$-axis). The violin plot shows stomatal density ($y$-axis, log-scale) as a function of \el. The width of the grey polygons indicates the density of data. Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles. Points outside the polygons are statistical outliers.} 
\label{fig:SD-light}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics{figures/figureS_violin.pdf}}
\caption{Most hydrophytes are hyperstomatous, having most stomata on the adaxial (`upper') surface (high $\mathrm{SD_{propAd}}$). The violin plot shows stomatal ratio as a function of Raunki\ae r lifeform. The width of the grey polygons indicates the density of data. Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles. Sample sizes per lifeform in the dataset are given above the upper plot margin. $\mathrm{SD_{ad}}$ and $\mathrm{SD_{total}}$ stand for the stomatal density on the adaxial surface and the total leaf surface (adaxial plus abaxial), respectively.}
\label{fig:violin}
\end{figure}

### 2017-07-12 ms/ms.tex
added ability to knit ms

% TASKS
% what's going on with monocots?
% in methods - mention stuff about Salisbury

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood. In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on 374 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item (return to this) key results: \el, growth form, and interaction are important
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
\end{itemize}

@@

I excluded data from hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. I also excluded C$_4$ (3 species) and CAM (2 species) plants.. I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data. The final dataset contained 374 species. The R code accompanying this paper documents these decisions with citations to the relevant literature \citep{Muir_dryad}.

@@

Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{4, 369}$ = 18.4, $P$ = 8.1 $\times10^{-14}$). Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \el~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}). Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals did not overlap zero (Fig.~\ref{fig:SRmultReg}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distibuted across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}, parametric bootstrap $P = 0.002$).

@@

Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio. The contributions of adaxial density, abaxial density, and their covariance are 1.14, 0.39, and -0.53, respectively. Recall that values can be greater than one for adaxial stomatal density and negative for the covariance when the latter value is positive. This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.

Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}). Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 8.8 $\times10^{-7}$ and 4.9 $\times10^{-5}$, respectively). However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2.1$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.2 versus 0.1). Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.22-fold change in adaxial stomatal density versus a 1.1-fold change in abaxial stomatal density. The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.1 $\times10^{-11}$). These results together imply that total stomatal density increases with \el, but the response is mediated mostly by adaxial stomatal density.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform shape stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ lifeform
  & 0.46
  & 0.068
  & 0.34
  & 12
  & -32.6
  & 89.2
  & 0 \\
    \el~+ lifeform
  & 0.47
  & 0.071
  & 0.32
  & 8
  & -39.8
  & 95.6
  & 6.3 \\
    \el
  & 0.65
  & 0.109
  & 0.26
  & 4
  & -58.9
  & 125.9
  & 36.6 \\
    lifeform
  & 0.34
  & 0.066
  & 0.15
  & 7
  & -79.2
  & 172.4
  & 83.2 \\
    1
  & 0.29
  & 0.068
  & 0
  & 3
  & -107.9
  & 221.7
  & 132.5 \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}

>>>>

% Notes from call with Rob
% van Tienderen - sensitivity approach better for evolution
% send list of species to rob

% TASKS
% what's going on with monocots?
% in methods - mention stuff about Salisbury
% leaf size

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood. In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item (return to this) key results: \el, growth form, and interaction are important
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
\end{itemize}

@@

I excluded data from hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. I also excluded C$_4$ (3 species) and CAM (2 species) plants.. I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data. The final dataset contained 372 species. The R code accompanying this paper documents these decisions with citations to the relevant literature \citep{Muir_dryad}.

@@

Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$). Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \el~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}). Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals did not overlap zero (Fig.~\ref{fig:SRmultReg}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distibuted across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).

@@

Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio. The contributions of adaxial density, abaxial density, and their covariance are 1.14, 0.38, and -0.53, respectively. Recall that values can be greater than one for adaxial stomatal density and negative for the covariance when the latter value is positive. This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.

Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}). Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 6.1 $\times10^{-7}$ and 2.9 $\times10^{-5}$, respectively). However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2.1$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.21 versus 0.1). Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.23-fold change in adaxial stomatal density versus a 1.1-fold change in abaxial stomatal density. The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.7 $\times10^{-11}$). These results together imply that total stomatal density increases with \el, but the response is mediated mostly by adaxial stomatal density.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform shape stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ lifeform
  & 0.46
  & 0.068
  & 0.34
  & 12
  & -33.3
  & 90.6
  & 0 \\
    \el~+ lifeform
  & 0.47
  & 0.072
  & 0.32
  & 8
  & -40.3
  & 96.5
  & 6 \\
    \el
  & 0.64
  & 0.108
  & 0.26
  & 4
  & -59.3
  & 126.6
  & 36.1 \\
    lifeform
  & 0.34
  & 0.067
  & 0.15
  & 7
  & -79.2
  & 172.4
  & 81.8 \\
    1
  & 0.29
  & 0.067
  & 0
  & 3
  & -107.5
  & 221
  & 130.5 \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}
